\documentclass{article}

\include{common}

\title{Programs and Methods}
\date{}

\begin{document}

\maketitle

\section{The Anatomy of a Java Program}


It is customary for a programmer's first program in a new language to be ``Hello, World.''  Here's our \href{\code/basics/HelloWorld.java}{{\tt HelloWorld.java}} program:
\begin{lstlisting}[language=Java]
public class HelloWorld {
    public static void main(String[] args) {
        System.out.println("Hello, world!");
    }
}
\end{lstlisting}

\begin{itemize}
\item The first line declares our {\tt HelloWorld} class.  {\tt class} is the syntax for declaring a class, and prepending with the {\tt public} modifer means the class will be visible outside {\tt HelloWorld}'s package.
\item Because we didn't declare a package explicitly, {\tt HelloWorld} is in the {\it default} package.  More on packages in a future lectrue.
\item The code between the curly braces, {\tt \{ ... \}} define the contents of the {\tt HelloWorld} class, in this case a single method, {\tt main}
\end{itemize}


\section{{\tt public static void main(String[] args)}}


In order to make a class executable with the {\tt java} command, it must have a main method:
\begin{lstlisting}[language=Java]
public static void main(String[] args) { ... }
\end{lstlisting}

\begin{itemize}
\item The {\tt public} modifier means we can call this method from outside the class.
\item The {\tt static} modifer means the method can be called without instantiating an object of the class.  Static methods (and variables) are sometimes called {\it class} methods.
\item {\tt void} is the return type.  In particular, main returns nothing.  Sometimes such subprograms are called {\it procedures} and distinguished from {\it functions}, which return values.
\item After the method name, {\tt main}, comes the parameter list.  {\tt main} takes a single parameter of type {\tt String[]} - an array of {\tt String}s.  {\tt args} is the name of the parameter, which we can refer to within the body of {\tt main}
\end{itemize}


\section{Methods}

The {\tt main} method is a special method that is used as the entry point for a Java program.  We can define other methods as well.  Consider this method from \href{\code/basics/NameParser.java}{NameParser.java}:

\begin{lstlisting}[language=Java]
public static String extractLastName(String name) {
    int commaPos = name.indexOf(",");
    int len = name.length();
    String lastName = name.substring(0, commaPos).trim();
    return lastName;
}
\end{lstlisting}
Similar to our {\tt main} method but:
\begin{itemize}
\item {\tt return}s a {\tt String} value
\item takes a single parameter of type {\tt String}
\end{itemize}


\section{Method Parameters}

In this method:
\begin{lstlisting}[language=Java]
public static String extractLastName(String name) {
    int commaPos = name.indexOf(",");
    int len = name.length();
    String lastName = name.substring(0, commaPos).trim();
    return lastName;
}
\end{lstlisting}
{\tt name} is a {\it parameter} (or {\it formal parameter}), a local scope variable within the {\tt extractLastName} method.  It is bound to a value when the method is called.  In the statement:

\begin{lstlisting}[language=Java]
String lastName = extractLastName(fullName);
\end{lstlisting}
the right-hand side, {\tt extractLastName(fullName)}, is a {\it method invocation} (or {\it method call}).  We say that {\tt fullName} is the {\it argument} (or {\it actual parameter}) to this invocation of the {\tt extractLastName} method.


\section{Local Variables}

Method parameters and variables declared inside the method are local to the method, invisible outside the method.  Local variables "shadow" variables of the same name in an enclosing scope.

\begin{lstlisting}[language=Java]
public class Methods {

    // Note that static variables cannot appear inside methods
    static String message = "Global message.";
    static int a = 5;

    public static int add(int a, int b) {
        String message = "Adding " + a + " and " + b;
        System.out.println(message);
        return a + b;
    }
...
\end{lstlisting}
In the {\tt add} method, the parameter {\tt a} shadows the {\tt static} variable {\tt a}, and the local variable {\tt message} shadows the {\tt static} variable {\tt message}.


\section{Methods as Expressions}

Methods that return values are expressions which can be used anywhere a value of the method's return type can be used.  Given:

\begin{lstlisting}[language=Java]
public static int add(int a, int b) { ... }
\end{lstlisting}
which returns an {\tt int}, this:
\begin{lstlisting}[language=Java]
x + (x + y)
\end{lstlisting}
is equivalent to this:
\begin{lstlisting}[language=Java]
x + add(x, y)
\end{lstlisting}
See \href{\code/basics/Methods.java}{Methods.java}.



\section{Closing Thoughts}

Methods are subprograms with
\begin{itemize}
\item input (parameters),
\item processing (a sequence of statements), and
\item output (return value).
\end{itemize}

Methods are a powerful form of procedural abstraction, another step in the building of complex programs from simple parts.


\end{document}
